Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies

### 2014-02-06 paper/Intuition.tex
adding tildas

Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
follows.
%From the transitions of Figure \ref{trans}, a store may happen only
%when a leaf cache is in \Mo{} state for that
%address (Invariant~\textit{processReq}).
Leaf caches cannot be ancestors of any other node, so by Invariant
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
leaf caches will be in \In{} state, blocking simultaneous
load and store requests (because of Invariant \ref{processReq}) for the same
address at the same time at different leaf caches. Thus, we have proved
Definition~\fullref{uniqStTimes}.

It is easy to see that Definition \fullref{uniqRespLabels}, Definition
\fullref{localOrdering}, and Definition \fullref{allPrevious} directly follow
from Invariant \fullref{noReqAgain}.  As leaf caches have no children, one in
at least \Sh{} state for address $a$ is \clean{} for that address. Since
processing a load request requires the cache to be at least in the \Sh{} state,
its response, which supplies the data for that address present in the cache,
supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants
\ref{latestValue} to \ref{noReqAgain}.

These arguments are fleshed out in a machine-checked Coq proof that the invariants
imply store atomicity.  Our proof may be reused for any protocol (i.e., alternative
choice of rules like in Figure \ref{trans}) that satisfies the invariants,
so that the only new verification effort is in proving the invariants.

@@

While Invariants \fullref{processReq} and \fullref{noReqAgain} can be seen
directly from the MSI transitions, correctness of Invariants
\fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
decomposition. We present 4 more invariants that are helpful in the proof.

@@

Invariants \fullref{conservative} and \fullref{localCompatibility1} defines a
$\ge$ relation between the states of a parent and its child for any address.
Since the ancestor relation is the reflexive-transitive closure of the parent
relation, the $\ge$ relation is preserved between the states of two caches $p$
and $c$ where $p$ is the ancestor of $c$.

Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
lowest common ancestor $x$ that is an ancestor of both $c_1$ and $c_2$. $x$
has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
$c_1$ and $c_2$ respectively. For any address $a$, $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible by 
Invariant~\fullref{localCompatibility2}. It follows that $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible since these values are smaller than the
respective directory values. From the previous paragraph about the relation
between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
are also compatible. This proves Invariant \fullref{nonAncestorCompatible}.

We prove Invariant \fullref{latestValue} by strong induction on time. Consider
two cases: (a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
(b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant
\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$ and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants cannot be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they cannot process any store requests.
Therefore, no cache save $c$ may have processed a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

If $c$ was not clean at time $t$, it must have received data from either its
parent or its child, and the sender is \clean{} by Invariant
\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
the latest value for the address. Again, because of
Invariant~\ref{noTransitWrite}, no store request can be processed when the data
is in transit, which means $c$ will have the latest value for address $a$ at
time $t+1$, establishing Invariant \fullref{latestValue}.

@@

Invariant \fullref{stateChange} ensures that a response message acts as a sync
between a parent's directory entry for a child and the state of the child, as
both the sender and the receiver change their respective directory entry or
state to the \myto-field of the response.

@@

Figure \ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

Using these three invariants (Invariants \ref{stateChange} to \ref{respFIFO}), it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure \ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state, and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section \ref{secondLevel} requires using
more properties of the MSI transistions of Figure \ref{trans}, though the argument is
straightforward. Analyzing a parent-child pair will suffice in proving Invariants
\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
\fullref{noTransitWrite} is similar to the proof of Invariant
\fullref{nonAncestorCompatible}.

>>>>

Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
follows.
%From the transitions of Figure \ref{trans}, a store may happen only
%when a leaf cache is in \Mo{} state for that
%address (Invariant~\textit{processReq}).
Leaf caches cannot be ancestors of any other node, so by Invariant~
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
leaf caches will be in \In{} state, blocking simultaneous
load and store requests (because of Invariant~\ref{processReq}) for the same
address at the same time at different leaf caches. Thus, we have proved
Definition~\fullref{uniqStTimes}.

It is easy to see that Definition~\fullref{uniqRespLabels}, Definition~
\fullref{localOrdering}, and Definition~\fullref{allPrevious} directly follow
from Invariant~\fullref{noReqAgain}.  As leaf caches have no children, one in
at least \Sh{} state for address $a$ is \clean{} for that address. Since
processing a load request requires the cache to be at least in the \Sh{} state,
its response, which supplies the data for that address present in the cache,
supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants~
\ref{latestValue} to \ref{noReqAgain}.

These arguments are fleshed out in a machine-checked Coq proof that the invariants
imply store atomicity.  Our proof may be reused for any protocol (i.e., alternative
choice of rules like in Figure~\ref{trans}) that satisfies the invariants,
so that the only new verification effort is in proving the invariants.

@@

While Invariants~\fullref{processReq} and \fullref{noReqAgain} can be seen
directly from the MSI transitions, correctness of Invariants~
\fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
decomposition. We present 4 more invariants that are helpful in the proof.

@@

Invariants~\fullref{conservative} and \fullref{localCompatibility1} defines a
$\ge$ relation between the states of a parent and its child for any address.
Since the ancestor relation is the reflexive-transitive closure of the parent
relation, the $\ge$ relation is preserved between the states of two caches $p$
and $c$ where $p$ is the ancestor of $c$.

Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
lowest common ancestor $x$ that is an ancestor of both $c_1$ and $c_2$. $x$
has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
$c_1$ and $c_2$ respectively. For any address $a$, $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible by 
Invariant~\fullref{localCompatibility2}. It follows that $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible since these values are smaller than the
respective directory values. From the previous paragraph about the relation
between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
are also compatible. This proves Invariant~\fullref{nonAncestorCompatible}.

We prove Invariant~\fullref{latestValue} by strong induction on time. Consider
two cases: (a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
(b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant~
\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$ and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants cannot be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they cannot process any store requests.
Therefore, no cache save $c$ may have processed a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

If $c$ was not clean at time $t$, it must have received data from either its
parent or its child, and the sender is \clean{} by Invariant~
\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
the latest value for the address. Again, because of
Invariant~\ref{noTransitWrite}, no store request can be processed when the data
is in transit, which means $c$ will have the latest value for address $a$ at
time $t+1$, establishing Invariant~\fullref{latestValue}.

@@

Invariant~\fullref{stateChange} ensures that a response message acts as a sync
between a parent's directory entry for a child and the state of the child, as
both the sender and the receiver change their respective directory entry or
state to the \myto-field of the response.

@@

Figure~\ref{resps} shows an example of the effect of Invariants~
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

Using these three invariants (Invariants~\ref{stateChange} to \ref{respFIFO}), it
is easy to see that Invariant~\fullref{conservative} holds. Intuitively, going
back to Figure~\ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state, and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant~\fullref{conservative} holds. Since Invariants~\ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section \ref{secondLevel} requires using
more properties of the MSI transistions of Figure~\ref{trans}, though the argument is
straightforward. Analyzing a parent-child pair will suffice in proving Invariants~
\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant~
\fullref{noTransitWrite} is similar to the proof of Invariant~
\fullref{nonAncestorCompatible}.

### 2014-02-06 paper/System.tex
maybe fixed another of adamc's comments?
adding tildas

Each node contains the fields mentioned in Figure \ref{hier}. We will
explain what the fields mean in the following discussion.

@@

\begin{figure}
\begin{subfigure}{6.8cm}
\begin{tabular}{|lp{5.8cm}|}
\hline
\multicolumn{2}{|c|}{A message in \cpReq}\\
\hline
\from: & Current state of child\\
\myto: & Desired (higher) state of child\\
\addr: & Address associated with the request\\
\hline
\hline
\multicolumn{2}{|c|}{A message in \cpResp}\\
\hline
\from: & Current state of child\\
\myto: & Downgraded state of child\\
\addr: & Address associated with the request\\
\data: & Data associated with the address, if necessary\\
\hline
\end{tabular}
\end{subfigure}
\begin{subfigure}{5.4cm}
\begin{tabular}{|lp{4.4cm}|}
\hline
\multicolumn{2}{|c|}{A message in \pc}\\
\hline
\typ: & \Req{} (for requests) and \Resp{} (for responses)\\
\myto: & Desired (lower) directory state of a child for \Req{} and upgraded
directory state of a child for \Resp{}\\
\addr: & Address associated with the request\\
\data: & Data associated with the address if necessary (for \Resp{} only)\\
\hline
\end{tabular}
\end{subfigure}
\caption{Data types for messages between caches
\adamc{Names here don't match those in the preceding text.}}
\label{format}
\end{figure}

@@

The behavior of the system is described via the atomic state transitions, which
are effectively a nondeterministic operational semantics. These transitions
have two parts: (a) the guard dictating when the transition can fire, and (b)
the action defining how the state of the system changes when the transition
happens. Logically, these transitions happen one at a time.  Figure~\ref{trans}
gives the atomic transitions for the MSI protocol.  The actions in the
transitions in Figure \ref{trans} show only the specific states that change,
and other state components are assumed to remain the same.  The notation ($x
\Leftarrow y$;) in a transition indicates setting of a specific state $x$ of
the overall system to value $y$.

@@

Figure~\ref{childside} shows the transitions that result in an upgrade of the
state of a cache. It involves a cache sending an upgrade request to the its
parent if the cache is not already waiting for some response for that address
from the parent (ChildSendReq). This request is received by the parent
(ParentRecvReq). If the directory entries for all of the other children are
\emph{compatible} with the upgrade request, and the parent's permission state has
at least the requested value, it sends an upgrade response to the cache, while
changing the appropriate directory state. The notation \compat{} used in
the ParentRecvReq transition of Figure \ref{trans} is defined as follows:
\begin{multline*}
\small
\compat(p, c, a, s) = \forall i \ne c. \;
\mylet s_i := p.\dstate[i][a] \myin\\
 (s = \Mo @-> s_i = \In) \wedge (s = \Sh @-> s_i = \Sh) \wedge (s = \In @-> s_i = \Mo)
\end{multline*}
Finally, the child cache receives the response from the parent (ChildRecvResp)
and sets its state while resetting its wait state if the response contains the
desired upgrade.

>>>>

Each node contains the fields mentioned in Figure~\ref{hier}. We will
explain what the fields mean in the following discussion.

@@

\begin{figure}
\begin{subfigure}{6.8cm}
\begin{tabular}{|lp{5.8cm}|}
\hline
\multicolumn{2}{|c|}{A message in \cpReq{} channel}\\
\hline
\from: & Current state of child\\
\myto: & Desired (higher) state of child\\
\addr: & Address associated with the request\\
\hline
\hline
\multicolumn{2}{|c|}{A message in \cpResp{} channel}\\
\hline
\from: & Current state of child\\
\myto: & Downgraded state of child\\
\addr: & Address associated with the request\\
\data: & Data associated with the address, if necessary\\
\hline
\end{tabular}
\end{subfigure}
\begin{subfigure}{5.4cm}
\begin{tabular}{|lp{4.4cm}|}
\hline
\multicolumn{2}{|c|}{A message in \pc{} channel}\\
\hline
\typ: & \Req{} (for requests) and \Resp{} (for responses)\\
\myto: & Desired (lower) directory state of a child for \Req{} and upgraded
directory state of a child for \Resp{}\\
\addr: & Address associated with the request\\
\data: & Data associated with the address if necessary (for \Resp{} only)\\
\hline
\end{tabular}
\end{subfigure}
\caption{Data types for messages between caches
\adamc{Names here don't match those in the preceding text.}}
\label{format}
\end{figure}

@@

The behavior of the system is described via the atomic state transitions, which
are effectively a nondeterministic operational semantics. These transitions
have two parts: (a) the guard dictating when the transition can fire, and (b)
the action defining how the state of the system changes when the transition
happens. Logically, these transitions happen one at a time.  Figure~\ref{trans}
gives the atomic transitions for the MSI protocol.  The actions in the
transitions in Figure~\ref{trans} show only the specific states that change,
and other state components are assumed to remain the same.  The notation ($x
\Leftarrow y$;) in a transition indicates setting of a specific state $x$ of
the overall system to value $y$.

@@

Figure~\ref{childside} shows the transitions that result in an upgrade of the
state of a cache. It involves a cache sending an upgrade request to the its
parent if the cache is not already waiting for some response for that address
from the parent (ChildSendReq). This request is received by the parent
(ParentRecvReq). If the directory entries for all of the other children are
\emph{compatible} with the upgrade request, and the parent's permission state has
at least the requested value, it sends an upgrade response to the cache, while
changing the appropriate directory state. The notation \compat{} used in
the ParentRecvReq transition of Figure~\ref{trans} is defined as follows:
\begin{multline*}
\small
\compat(p, c, a, s) = \forall i \ne c. \;
\mylet s_i := p.\dstate[i][a] \myin\\
 (s = \Mo @-> s_i = \In) \wedge (s = \Sh @-> s_i = \Sh) \wedge (s = \In @-> s_i = \Mo)
\end{multline*}
Finally, the child cache receives the response from the parent (ChildRecvResp)
and sets its state while resetting its wait state if the response contains the
desired upgrade.
